# Reproducible Research
#### Peer Assessment 1

```{r message=FALSE, warning=FALSE}
# Load some libraries
#library(sqldf)
#library(plyr)
library(ggplot2)
#library(lattice)
library(knitr)
library(markdown)

```


##1 - Loading and preprocessing the data

### 1.1 - Load the data
```{r}
# Extract the file activity.csv from the zip file, if it does not exist in the directory
if(!file.exists("activity.csv"))
    unzip("activity.zip")

# Read the file into memory
dtAll <- read.csv("activity.csv")
```


### 1.2 - Process and transform the data

```{r}
# Remove rows with NAs on the steps column
dtNNA <- as.data.frame(dtAll[!is.na(dtAll$steps), ])
```


## 2 - What is mean total number of steps taken per day?

### 2.1 - Total number of steps taken each day

```{r fig.align='center'}
# Aggregate by Date, using function sum to get the total number of steps, by day

x2 <- aggregate(steps ~ date, data=dtAll, FUN=function(x) sum(x, na.rm=F))

# Plot the total steps taken each day
barplot(x2$steps, plot=TRUE, main="Total number of steps taken each day", col="LightBlue", xlab="Date", ylab="Steps", names.arg=x2$date, space=0)
box()
```

### 2.2 - Report with the mean and median total number of steps taken per day
```{r warnnig=FALSE, message=FALSE}
y2 <- data.frame(cbind(sprintf("%.2f", mean(x2$steps)), sprintf("%.2f", median(x2$steps))))
names(y2) <- c("Mean", "Median")
print(y2)
```


## 3 - What is the average daily activity pattern?

### 3.1 - Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r fig.align='center'}
x3 <- aggregate(steps ~ interval, data=dtAll, FUN=function(x) mean(x, na.rm=TRUE))
plot(x3, type="l")
```

### 3.2 - Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r results="asis"}
y <- x3[x3$steps == max(x3$steps), ]
sprintf("The 5-minute interval with the maximum number of steps is %.0f, with the value %.2f", y$interval, max(y$steps))
```

## 4 - Imputing missing values

### 4.1 - Total number of missing values (NAs)

```{r}
sprintf("The total number of rows with NAs is: %.0f", sum(is.na(dtAll$steps)))
```

### 4.2 - Filling missing values (NAs)

```{r}
# Use the dataset 'x3', produced in point 3.1 with the average for every 5-minute across all days, excluding NAs, to fill dtALL NAs rows with new values

# Create a dataset with NAs rows only
a <- dtAll[is.na(dtAll$steps),]

# Merge with x3 to get the new value for the steps
b <- merge(a, x3, by="interval")

# Select only the columns of interest
c <- subset(b, select=c(steps.y, date, interval))
names(c) <- c("steps", "date", "interval")

# Create a data with all the row with values for the step
dtAllFill <- rbind(dtAll[!is.na(dtAll$steps),], c)
```

### 4.3 - Create a new dataset with NAs removed
This is the dataset **dtAllFill** created on point 4.2

### 4.4 - Make an histogram and report the mean and median for the new dataset
The code is similar to the one used on points 2.1 and 2.2


```{r fig.align='center'}
# Total number of steps taken each day
# Aggregate by Date, using function sum to get the total number of steps, by day

x4 <- aggregate(steps ~ date, data=dtAllFill, FUN=sum)

# Plot the total steps taken each day
barplot(x4$steps, plot=TRUE, main="Total number of steps taken each day", col="LightBlue", xlab="Date", ylab="Steps", names.arg=x4$date, space=0)
box()

### Report with the mean and median total number of steps taken per day
y4 <- data.frame(cbind(sprintf("%.2f", mean(x4$steps)), sprintf("%.2f", median(x4$steps))))
names(y4) <- c("Mean", "Median")
print(y4)
```

There are 2304 rows with the column steps with NAs. Changing the value of this column has no impact on the mean and the median changes a litle bit.

## 5 - Are there differences in activity patterns between weekdays and weekends?

### 5.1 - Create a new factor variable for **weekday** and **weekend**

```{r}
# Using the dataset dtAllFill that contains all the rows created on point 4.2
# Add the weekdayf factor column
dtAllFill <- cbind(dtAllFill, weekday=weekdays(as.Date(dtAllFill$date), abbreviate=TRUE), weekdayf = c("weekday", "weekend"))

# Categorize the days between weekdays and weekend
dtAllFill[dtAllFill$weekday %in% c("seg", "ter", "qua", "qui", "sex"), "weekdayf"] <- "weekday"
dtAllFill[dtAllFill$weekday %in% c("sáb", "dom"), "weekdayf"] <- "weekend"

# Create a dataset to plot by weekday and interval
x5 <- aggregate(steps ~ weekdayf + interval, data=subset(dtAllFill, select=c(steps, interval, weekdayf)), FUN=function(x) Mean=sprintf("%.2f", mean(x)))
```

### 5.2 - Plot the timeseries for weekday and weekend

```{r fig.align='center'}
par(mfrow=c(1, 2))
plot(x5[x5$weekdayf == "weekday", c("interval", "steps")], type="l", main="Weekday")
plot(x5[x5$weekdayf == "weekend", c("interval", "steps")], type="l", main="Weekend")
```
   